---
title: "Bonobo_steepness"
author: "Lucas Spicher"
date: "2025-01-13"
output: html_document
---
##Pan paniscus
Bonobo

1) De vries et al., 2006: 1 matrix of 7 bonobos but without infomartions in the paper
2) Fortunato 2009 (PhD thesis) : no access to the thesis file, no information
3) Furuichi 1997: it's all on the paper (sex & age)
4) Paoli & Palagi 2008: 2 matrices but no access to the paper
5) Surbeck et Hohmann 2013: 1 matrix and it's all on the paper (sex & age)
6) Vervaecke et al., 2000: 1 matrix and it's all on the paper (sex & age)

```{r}
library(EloSteepness.data)
library(EloSteepness)
library(EloRating)
```

I get the matrices from the EloSteepness.data package
```{r}
bono1 <- empirical_matlist$`devries2006_dom_1_@_mix_@_step00`
bono2 <- empirical_matlist$`fortunato2009_dom_1_@_by_dyad_@_step00`
bono3 <- empirical_matlist$`furuichi1997_dom_1_@_mix_@_step00`
bono4 <- empirical_matlist$`paoli2008_dom_1_@_mix_@_step00`
bono5 <- empirical_matlist$`paoli2008_dom_2_@_by_interaction_@_step00`
bono6 <- empirical_matlist$`surbeck2013_dom_1_@_mix_@_step00`
bono7 <- empirical_matlist$`vervaecke2000_dom_1_@_by_interaction_@_step00`
```

For each steep_bono, we obtain a summary with median steepness and the 89% CI
depending on your computer’s performance that step may take anything between less than a minute and several minutes (or even more. . . )

The number of randomizations is set in the following way, unless a specific number is provided. If there are more than 500 observed interactions, n_rand = 5. If there are less than 100 interactions, n_rand = 50. In the remaining cases, n_rand = 20.
```{r}
steep_bono1 <- elo_steepness_from_matrix(mat = bono1, 
                                         n_rand = 20, 
                                         refresh = 0, 
                                         cores = 2, 
                                         iter = 2000, 
                                         seed = 1)

summary(steep_bono1)

steep_bono2 <- elo_steepness_from_matrix(mat = bono2, 
                                         n_rand = 20, 
                                         refresh = 0, 
                                         cores = 2, 
                                         iter = 2000, 
                                         seed = 1)

summary(steep_bono2)

steep_bono3 <- elo_steepness_from_matrix(mat = bono3, 
                                         n_rand = 50, 
                                         refresh = 0, 
                                         cores = 2, 
                                         iter = 2000, 
                                         seed = 1)

summary(steep_bono3)

steep_bono4 <- elo_steepness_from_matrix(mat = bono4, 
                                         n_rand = 20, 
                                         refresh = 0, 
                                         cores = 2, 
                                         iter = 2000, 
                                         seed = 1)

summary(steep_bono4)

steep_bono5 <- elo_steepness_from_matrix(mat = bono5, 
                                         n_rand = 20, 
                                         refresh = 0, 
                                         cores = 2, 
                                         iter = 2000, 
                                         seed = 1)

summary(steep_bono5)

steep_bono6 <- elo_steepness_from_matrix(mat = bono6, 
                                         n_rand = 5, 
                                         refresh = 0, 
                                         cores = 2, 
                                         iter = 2000, 
                                         seed = 1)

summary(steep_bono6)

steep_bono7 <- elo_steepness_from_matrix(mat = bono7, 
                                         n_rand = 5, 
                                         refresh = 0, 
                                         cores = 2, 
                                         iter = 2000, 
                                         seed = 1)

summary(steep_bono7)


```


Posterior distribution of the steepness values.

```{r}
plot_steepness(steep_bono1)
plot_steepness(steep_bono2)
plot_steepness(steep_bono3)
plot_steepness(steep_bono4)
plot_steepness(steep_bono5)
plot_steepness(steep_bono6)
plot_steepness(steep_bono7)

#The yellow area gives a visual summary of the posterior combined across all randomizations. The finer gray lines represent the posteriors for each individual randomization. By default, the plot also provides some summary statistics, which in this case shows that mean and median of the posterior correspond to very similar values.
```


individual scores (‘ranks’) & steepness regression
1) the first line defines the colour vector (need to be the same that the group size) 
2) plot_scores (individua sores) :  on which the steepness calculation is based, i.e. the dominance ‘ranks’ of individuals. 
3) plot_steepness_regression : combine steepness with the individual scores.
```{r}
my_colors7 <- hcl.colors(n = 7, palette = "zissou1", alpha = 0.7) 
plot_scores(steep_bono1, color = my_colors7)
plot_steepness_regression(steep_bono1, color = my_colors7, width_fac = 0.3)
# width_fac= argument. This is just a visual adjustment of how ‘high’ the highest individual posterior extends (along the horizontal axis).

my_colors9 <- hcl.colors(n = 9, palette = "zissou1", alpha = 0.7) 
plot_scores(steep_bono2, color = my_colors9)
plot_steepness_regression(steep_bono2, color = my_colors9, width_fac = 0.3)

plot_scores(steep_bono3, color = my_colors9)
plot_steepness_regression(steep_bono3, color = my_colors9, width_fac = 0.3)

my_colors8 <- hcl.colors(n = 8, palette = "zissou1", alpha = 0.7) 
plot_scores(steep_bono4, color = my_colors8)
plot_steepness_regression(steep_bono4, color = my_colors8, width_fac = 0.3)

my_colors6 <- hcl.colors(n = 6, palette = "zissou1", alpha = 0.7) 
plot_scores(steep_bono5, color = my_colors6)
plot_steepness_regression(steep_bono5, color = my_colors6, width_fac = 0.3)

my_colors21 <- hcl.colors(n = 21, palette = "zissou1", alpha = 0.7) 
plot_scores(steep_bono6, color = my_colors21)
plot_steepness_regression(steep_bono6, color = my_colors21, width_fac = 0.3)

plot_scores(steep_bono7, color = my_colors6)
plot_steepness_regression(steep_bono7, color = my_colors6, width_fac = 0.3)

```

My PC only has 2 cores, so each steep_ can take a long time to run, so I save the R object on my desktop for each species.
```{r}
save.image(file = "~/Desktop/P.paniscus.RData")
```

Then I create the Excel file, which I also save on my desktop 
```{r}
library(openxlsx)

# Converting the matrix into a data frame
bono1df <- as.data.frame(bono1)
bono2df <- as.data.frame(bono2)
bono3df <- as.data.frame(bono3)
bono4df <- as.data.frame(bono4)
bono5df <- as.data.frame(bono5)
bono6df <- as.data.frame(bono6)
bono7df <- as.data.frame(bono7)

# Add the names of individuals (rownames) as a column
bono1df <- cbind(Individus = rownames(bono1df), bono1df)
bono2df <- cbind(Individus = rownames(bono2df), bono2df)
bono3df <- cbind(Individus = rownames(bono3df), bono3df)
bono4df <- cbind(Individus = rownames(bono4df), bono4df)
bono5df <- cbind(Individus = rownames(bono5df), bono5df)
bono6df <- cbind(Individus = rownames(bono6df), bono6df)
bono7df <- cbind(Individus = rownames(bono7df), bono7df)

# Create an Excel file
wbbono <- createWorkbook()

# Add data frames as sheets
addWorksheet(wbbono, "deVries_etal2006")
writeData(wbbono, sheet = "deVries_etal2006", bono1df)

addWorksheet(wbbono, "Fortunato2009")
writeData(wbbono, sheet = "Fortunato2009", bono2df)

addWorksheet(wbbono, "Furuichi1997")
writeData(wbbono, sheet = "Furuichi1997", bono3df)

addWorksheet(wbbono, "Paoli&Palagi_1")
writeData(wbbono, sheet = "Paoli&Palagi_1", bono4df)

addWorksheet(wbbono, "Paoli&Palagi_2")
writeData(wbbono, sheet = "Paoli&Palagi_2", bono5df)

addWorksheet(wbbono, "Surbeck&Hohmann2013")
writeData(wbbono, sheet = "Surbeck&Hohmann2013", bono6df)

addWorksheet(wbbono, "Vervaecke_etal2000")
writeData(wbbono, sheet = "Vervaecke_etal2000", bono7df)

# Save file to desktop
desktop_path <- "~/Desktop/matrices_pan_paniscus.xlsx" 

saveWorkbook(wbbono, file = desktop_path, overwrite = TRUE)
```

voila
###########################################################################

```{r}

library(dplyr)

# For the Surbeck 2013
bono6males<-c("ca","ti","ja","da","be")
bono6females<-c("ha","ri","ma","ev","ir","pa","zo","ol","gw","su")
bono6_malehierarchy<-empirical_raw_data$surbeck2013_dom_1[rownames(empirical_raw_data$surbeck2013_dom_1) %in% bono6males,colnames(empirical_raw_data$surbeck2013_dom_1) %in% bono6males]
bono6_femalehierarchy<-empirical_raw_data$surbeck2013_dom_1[rownames(empirical_raw_data$surbeck2013_dom_1) %in% bono6females,colnames(empirical_raw_data$surbeck2013_dom_1) %in% bono6females]


# The number of randomizations is set in the following way, unless a specific number is provided. If there are more than 500 observed interactions, n_rand = 5. If there are less than 100 interactions, n_rand = 50. In the remaining cases, n_rand = 20.
steep_bono6_males <- elo_steepness_from_matrix(mat = bono6_malehierarchy, 
                                         n_rand = case_when(sum(bono6_malehierarchy)>0 ~ 20,sum(bono6_malehierarchy)>500 ~ 5,sum(bono6_malehierarchy)<100 ~ 50  ), 
                                         refresh = 0, 
                                         cores = 2, 
                                         iter = 2000, 
                                         seed = 1)

steepness_bono6_males <-median(steep_bono6_males$steepness)


steep_bono6_females <- elo_steepness_from_matrix(mat = bono6_femalehierarchy, 
                                         n_rand = case_when(sum(bono6_femalehierarchy)>0 ~ 20,sum(bono6_femalehierarchy)>500 ~ 5,sum(bono6_femalehierarchy)<100 ~ 50  ), 
                                         refresh = 0, 
                                         cores = 2, 
                                         iter = 2000, 
                                         seed = 1)

steepness_bono6_females <-median(steep_bono6_females$steepness)

```