---
title: "TREE"
output: html_document
date: "2026-04-08"
---

We first load the required packages
```{r}
library(ggtree)
library(ggplot2)
library(dplyr)
library(stringr)
library(tidytree)
```

# Load the data
```{r}
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
data <- read.csv("Comparative_primate_steepness_values.csv")

#and prepare it
data_alt <- data %>%
  group_by(species, sex) %>%
  summarise(
    mean_steepness = mean(steepness, na.rm = TRUE),
    n_studies = n())
```


Our initial tree : 
```{r}
p <- ggtree(mtree) + theme_tree2()

data_N <- data %>%
  group_by(species) %>%
  summarise(Nmatrices = n(), .groups = "drop") %>%  # n() = nombre de lignes par espèce
  mutate(
    species_label = str_replace_all(species, "_", " "),
    species_label = str_to_sentence(species_label)
  )
p2 <- p %<+% data_N

p2 +
  geom_tiplab(aes(label = species_label), size = 4, offset = 4, align = TRUE,fontface = "italic") +
  geom_tippoint(aes(size = Nmatrices), shape = 21, fill = "steelblue", alpha = 0.7) +
  scale_size_area(
    max_size = 6,
    name = "N matrices",
    breaks = round(seq(min(data_N$Nmatrices), max(data_N$Nmatrices), length.out = 4))
  ) +
  coord_cartesian(xlim = c(0, 75), clip = "off")

p2 +
  geom_hilight(node = 70, fill = "#8da0cb", alpha = 0.2) +
  geom_hilight(node = 41, fill = "#66c2a5", alpha = 0.2) +
  geom_hilight(node = 67, fill = "#fc8d62", alpha = 0.2) +
  
  geom_tippoint(aes(size = Nmatrices), shape = 21, fill = "steelblue", alpha = 0.7) +
  
  geom_tiplab(aes(label = species_label),
              size = 3.5, offset = 4, align = TRUE,
              fontface = "italic") +
  
  geom_cladelabel(node = 70, label = "Platyrrhini",
                  align = TRUE,
                  offset = 25,
                  offset.text = 3,
                  fontsize = 4,
                  barsize = 1) +
  
  geom_cladelabel(node = 41, label = "Cercopithecoidea",
                  align = TRUE,
                  offset = 25,
                  offset.text = 3,
                  fontsize = 4,
                  barsize = 1) +
  
  geom_cladelabel(node = 67, label = "Hominoidea",
                  align = TRUE,
                  offset = 25,
                  offset.text = 3,
                  fontsize = 4,
                  barsize = 1) +
  scale_size_area(
    max_size = 6,
    name = "N matrices",
    breaks = round(seq(min(data_N$Nmatrices),
                       max(data_N$Nmatrices),
                       length.out = 4))
  ) +
  
  coord_cartesian(xlim = c(0, 95), clip = "off")

```

Phylogeny: good idea to include a phylogeny. The one you included in the manuscript works. The alternative would be to plot the average steepness of females and of males. That could also be used to illustrate the points for the within-species analysis, that on average the steepness within the sexes is not biased in a particular direction.

packages
```{r}
library(ggtree)
library(ggplot2)
library(dplyr)
library(stringr)
library(tidytree)
```

Data
```{r}
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
data <- read.csv("Comparative_primate_steepness_values.csv")

#phylo
specieslist<-as.data.frame(matrix(unique(data$species),ncol=1,nrow=length(unique(data$species))))
colnames(specieslist)<-"species"
rownames(specieslist)<-specieslist$species


phylogeny <- read.nexus("Upham2019MammalPhylogeny.nex")

# check whether the species in our data are the same as the species in the phylogeny
speciesmatching<-name.check(phylogeny,specieslist)
# check which species from our data are not in the phylogeny
speciesmatching$data_not_tree

specieslist<-as.data.frame(matrix(unique(data$species),ncol=1,nrow=length(unique(data$species))))
colnames(specieslist)<-"species"
rownames(specieslist)<-specieslist$species

# and reduce the phylogeny to only include the species for which we have data
speciesmatching<-name.check(phylogeny,specieslist)
mtree<-drop.tip(phylogeny,speciesmatching$tree_not_data)
```

Preparing the data
```{r}
#Nombre de matrices par espèce
data_N <- data %>%
  group_by(species) %>%
  summarise(Nmatrices = n(), .groups = "drop") %>%
  mutate(
    species_label = str_replace_all(species, "_", " "),
    species_label = str_to_sentence(species_label)
  )

#average steepness moyenes per espèce and sex
data_steep <- data %>%
  group_by(species, sex) %>%
  summarise(
    mean_steepness = mean(steepness, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    sex = recode(sex,
                 "M" = "males",
                 "F" = "females",
                 "male" = "males",
                 "female" = "females",
                 "males" = "males",
                 "females" = "females")
  )

data_steep_wide <- data_steep %>%
  pivot_wider(
    names_from = sex,
    values_from = mean_steepness
  )


```

Initial tree
```{r}
p <- ggtree(mtree) + theme_tree2()
p2 <- p %<+% data_N

#vertical spacing
y_expand <- 2
p2$data$y <- p2$data$y * y_expand
```

Extract tips positions
```{r}
tip_positions <- p2$data %>%
  filter(isTip) %>%
  select(label, x, y) %>%
  rename(species = label)

tree_max_x <- max(tip_positions$x)
```

Build values positions 
```{r}
text_data <- data_steep_wide %>%
  left_join(tip_positions, by = "species") %>%
  mutate(
    x_anchor  = tree_max_x + 27,
    x_male    = x_anchor,
    x_female  = x_anchor + 4.8
  )
```

legend
```{r}
clade_legend <- data.frame(
  clade = factor(c("Platyrrhini", "Cercopithecoidea", "Hominoidea"),
                 levels = c("Platyrrhini", "Cercopithecoidea", "Hominoidea")),
  x = NA, y = NA
)

sex_legend <- data.frame(
  sex = factor(c("males", "females"), levels = c("males", "females")),
  x = 1,
  y = 1,
  label = "0.88"
)
```

plot
```{r}
p_final <- p2 +
  geom_hilight(node = 70, fill = "#8da0cb", alpha = 0.18) +
  geom_hilight(node = 41, fill = "#66c2a5", alpha = 0.18) +
  geom_hilight(node = 67, fill = "#fc8d62", alpha = 0.18) +

  geom_tippoint(aes(size = Nmatrices),
                shape = 21, fill = "steelblue", alpha = 0.7) +

  geom_tiplab(aes(label = species_label),
              size = 3.8,
              offset = 4,
              align = TRUE,
              linetype = "dotted",
              fontface = "italic") +

 # males values
geom_text(
  data = (text_data %>% filter(!is.na(males)) %>% mutate(sex = "males")),
  aes(x = x_male, y = y, label = sprintf("%.2f", males), color = sex),
  inherit.aes = FALSE,
  hjust = 0,
  size = 3.5,
  fontface = "bold"
) +

# females values
geom_text(
  data = (text_data %>% filter(!is.na(females)) %>% mutate(sex = "females")),
  aes(x = x_female, y = y, label = sprintf("%.2f", females), color = sex),
  inherit.aes = FALSE,
  hjust = 0,
  size = 3.5,
  fontface = "bold"
) +

  # average steepness text
  annotate(
    "text",
    x = text_data$x_anchor[1] + 3,
    y = max(text_data$y) + 3.5,
    label = "Average steepness",
    fontface = "bold",
    size = 4
  ) +

  # clade legend
  geom_point(
    data = clade_legend,
    aes(x = x, y = y, fill = clade),
    shape = 22,
    size = 4,
    inherit.aes = FALSE
  ) +
  
  scale_fill_manual(
    values = c(
      "Platyrrhini" = "#8da0cb",
      "Cercopithecoidea" = "#66c2a5",
      "Hominoidea" = "#fc8d62"
    ),
    name = "Clade"
  ) +
  
  scale_color_manual(
    values = c("males" = "#FDE725FF", "females" = "#443A83FF"),
    name = NULL
  ) +

  scale_size_area(
    max_size = 6,
    name = "N matrices",
    breaks = round(seq(min(data_N$Nmatrices),
                       max(data_N$Nmatrices),
                       length.out = 4))
  ) +

  guides(
    fill = guide_legend(order = 1, override.aes = list(shape = 22, size = 5)),
    color = guide_legend(
      override.aes = list(
        label = c("0.88", "0.88"),
        shape = NA
      )
    ),
    size = guide_legend(order = 3)
  ) +

  coord_cartesian(
    xlim = c(0, max(text_data$x_female) + 5),
    clip = "off"
  ) +
  theme_tree2() +
  theme(
    legend.position = "right",
    plot.margin = margin(20, 70, 20, 10)
  )

p_final
```

We want to investogate a bit more the steepness in the 3 phylogenetic group

Descriptive about the phylogenetic group
```{r}
data$clade <- case_when(
  data$species %in% c("Pan_troglodytes", "Pan_paniscus", "Gorilla_beringei", "Gorilla_gorilla") ~ "Hominoidea",
  data$species %in% c("Cebus_capucinus", "Cebus_apella", "Saimiri_sciureus", "Alouatta_palliata") ~ "Platyrrhini",
  TRUE ~ "Cercopithecoidea" 
)

clade_summary <- data %>%
  group_by(clade) %>%
  summarise(
    N_matrices = n(),
    N_species = n_distinct(species),
    mean_steepness = round(mean(steepness, na.rm=T), 3),
    sd_steepness = round(sd(steepness, na.rm=T), 3),
    range_steepness = paste0(round(range(steepness, na.rm=T), 3), collapse = " - "),
    .groups = 'drop'
  )

clade_sex_summary <- data %>%
  group_by(clade, sex) %>%
  summarise(
    N_matrices = n(),
    N_species = n_distinct(species),
    mean_steepness = round(mean(steepness, na.rm=T), 3),
    .groups = 'drop'
  )
```

Put relevant informations in a dataframe
```{r}
clade_stats <- data %>%
  group_by(clade) %>%
  summarise(
    N_matrices = n(),
    N_species = n_distinct(species),
    
    # STEEPNESS
    median_steepness = round(median(steepness, na.rm=T), 3),
    mean_steepness = round(mean(steepness, na.rm=T), 3),
    ci_lower_steep = round(quantile(steepness, 0.10, na.rm=T), 3),
    ci_upper_steep = round(quantile(steepness, 0.90, na.rm=T), 3),
    
    # LINEARITY (h_index)
    median_linearity = round(median(h_index, na.rm=T), 3),
    mean_linearity = round(mean(h_index, na.rm=T), 3),
    ci_lower_lin = round(quantile(h_index, 0.10, na.rm=T), 3),
    ci_upper_lin = round(quantile(h_index, 0.90, na.rm=T), 3),
    
    .groups = 'drop'
  )
print(clade_stats)
```

```{r}
label_map <- c("Cercopithecoidea" = "Cercopithecoidea", 
               "Hominoidea" = "Hominoidea", 
               "Platyrrhini" = "Platyrrhini")
```


PLOT the median steepness and linearity in relation to the 3 group
```{r}
#plot steepness median
p_clades_steep_med <- ggplot() +
  geom_jitter(data = data,
              aes(x = clade, y = steepness, color = clade),
              width = 0.25, size = 5, alpha = 0.6, stroke = 0) +
  # Errorbar CI median
  geom_errorbar(data = clade_stats,
                aes(x = clade, ymin = ci_lower_steep, ymax = ci_upper_steep),
                width = 0.25, size = 0.5, color = "black") +
  geom_point(data = clade_stats,
             aes(x = clade, y = median_steepness),
             size = 5, shape = 21, fill = "white", color = "black", stroke = 2) +
  scale_color_manual(values = c("#66c2a5", "#fc8d62", "#8da0cb")) + 
  scale_x_discrete(labels = label_map) +
  labs(x = NULL, y = "Hierarchy Steepness") +
  ylim(0, 1) +
  theme_minimal(base_size = 16) +
  theme(axis.title = element_text(face = "bold", size = 16),
        axis.text = element_text(size = 10),
        legend.position = "none")


p_clades_lin_med <- ggplot() +
  geom_jitter(data = data,
              aes(x = clade, y = h_index, color = clade),
              width = 0.25, size = 5, alpha = 0.6, stroke = 0) +
  # Errorbar CI median
  geom_errorbar(data = clade_stats,
                aes(x = clade, ymin = ci_lower_lin, ymax = ci_upper_lin),
                width = 0.25, size = 0.5, color = "black") +
  geom_point(data = clade_stats,
             aes(x = clade, y = median_linearity),
             size = 5, shape = 21, fill = "white", color = "black", stroke = 2) +
  scale_color_manual(values = c("#66c2a5", "#fc8d62", "#8da0cb")) + 
  labs(x = "Clade", y = "Hierarchy Linearity") +
  ylim(0, 1) +
  theme_minimal(base_size = 16) +
  theme(
    axis.title = element_text(face = "bold", size = 16),
    axis.text = element_text(size = 10),
    legend.position = "none",
    panel.grid.minor = element_blank()
  )

# Merge them
x.grob_1 <- textGrob("Clade", gp = gpar(fontsize = 16, fontface = "bold"))
grid.arrange(p_clades_steep_med + xlab(NULL), p_clades_lin_med + xlab(NULL),
             ncol = 2, bottom = x.grob_1)
```


PLOT the mean steepness and linearity in relation to the 3 group
```{r}
#plot steepness mean
p_clades_steep_mean <- ggplot() +
  geom_jitter(data = data,
              aes(x = clade, y = steepness, color = clade),
              width = 0.25, size = 5, alpha = 0.6, stroke = 0) +
  # Errorbar CI mean
  geom_errorbar(data = clade_stats,
                aes(x = clade, ymin = ci_lower_steep, ymax = ci_upper_steep),
                width = 0.25, size = 0.5, color = "black") +
  geom_point(data = clade_stats,
             aes(x = clade, y = mean_steepness),
             size = 5, shape = 21, fill = "white", color = "black", stroke = 2) +
  scale_color_manual(values = c("#66c2a5", "#fc8d62", "#8da0cb")) + 
  scale_x_discrete(labels = label_map) +
  labs(x = NULL, y = "Hierarchy Steepness") +
  ylim(0, 1) +
  theme_minimal(base_size = 16) +
  theme(axis.title = element_text(face = "bold", size = 16),
        axis.text = element_text(size = 10),
        legend.position = "none")


p_clades_lin_med <- ggplot() +
  geom_jitter(data = data,
              aes(x = clade, y = h_index, color = clade),
              width = 0.25, size = 5, alpha = 0.6, stroke = 0) +
  # Errorbar CI mean
  geom_errorbar(data = clade_stats,
                aes(x = clade, ymin = ci_lower_lin, ymax = ci_upper_lin),
                width = 0.25, size = 0.5, color = "black") +
  geom_point(data = clade_stats,
             aes(x = clade, y = mean_linearity),
             size = 5, shape = 21, fill = "white", color = "black", stroke = 2) +
  scale_color_manual(values = c("#66c2a5", "#fc8d62", "#8da0cb")) + 
  labs(x = "Clade", y = "Hierarchy Linearity") +
  ylim(0, 1) +
  theme_minimal(base_size = 16) +
  theme(
    axis.title = element_text(face = "bold", size = 16),
    axis.text = element_text(size = 10),
    legend.position = "none",
    panel.grid.minor = element_blank()
  )

# Merge them
x.grob_1 <- textGrob("Clade", gp = gpar(fontsize = 16, fontface = "bold"))
grid.arrange(p_clades_steep_med + xlab(NULL), p_clades_lin_med + xlab(NULL),
             ncol = 2, bottom = x.grob_1)
```


###############################################################################











