---
title: "TREE"
output: html_document
date: "2026-04-08"
---

We first load the required packages
```{r}
library(ggtree)
library(ggplot2)
library(dplyr)
library(stringr)
library(tidytree)
```

# Load the data
```{r}
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
data <- read.csv("Comparative_primate_steepness_values.csv")

#and prepare it
data_alt <- data %>%
  group_by(species, sex) %>%
  summarise(
    mean_steepness = mean(steepness, na.rm = TRUE),
    n_studies = n())
```


Our initial tree : 
```{r}
p <- ggtree(mtree) + theme_tree2()

data_N <- data %>%
  group_by(species) %>%
  summarise(Nmatrices = n(), .groups = "drop") %>%  # n() = nombre de lignes par espèce
  mutate(
    species_label = str_replace_all(species, "_", " "),
    species_label = str_to_sentence(species_label)
  )
p2 <- p %<+% data_N

p2 +
  geom_tiplab(aes(label = species_label), size = 4, offset = 4, align = TRUE,fontface = "italic") +
  geom_tippoint(aes(size = Nmatrices), shape = 21, fill = "steelblue", alpha = 0.7) +
  scale_size_area(
    max_size = 6,
    name = "N matrices",
    breaks = round(seq(min(data_N$Nmatrices), max(data_N$Nmatrices), length.out = 4))
  ) +
  coord_cartesian(xlim = c(0, 75), clip = "off")

p2 +
  geom_hilight(node = 70, fill = "#8da0cb", alpha = 0.2) +
  geom_hilight(node = 41, fill = "#66c2a5", alpha = 0.2) +
  geom_hilight(node = 67, fill = "#fc8d62", alpha = 0.2) +
  
  geom_tippoint(aes(size = Nmatrices), shape = 21, fill = "steelblue", alpha = 0.7) +
  
  geom_tiplab(aes(label = species_label),
              size = 3.5, offset = 4, align = TRUE,
              fontface = "italic") +
  
  geom_cladelabel(node = 70, label = "Platyrrhini",
                  align = TRUE,
                  offset = 25,
                  offset.text = 3,
                  fontsize = 4,
                  barsize = 1) +
  
  geom_cladelabel(node = 41, label = "Cercopithecoidea",
                  align = TRUE,
                  offset = 25,
                  offset.text = 3,
                  fontsize = 4,
                  barsize = 1) +
  
  geom_cladelabel(node = 67, label = "Hominoidea",
                  align = TRUE,
                  offset = 25,
                  offset.text = 3,
                  fontsize = 4,
                  barsize = 1) +
  scale_size_area(
    max_size = 6,
    name = "N matrices",
    breaks = round(seq(min(data_N$Nmatrices),
                       max(data_N$Nmatrices),
                       length.out = 4))
  ) +
  
  coord_cartesian(xlim = c(0, 95), clip = "off")

```

Phylogeny: good idea to include a phylogeny. The one you included in the manuscript works. The alternative would be to plot the average steepness of females and of males. That could also be used to illustrate the points for the within-species analysis, that on average the steepness within the sexes is not biased in a particular direction.

packages
```{r}
library(ggtree)
library(ggplot2)
library(dplyr)
library(stringr)
library(tidytree)
```

Data
```{r}
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
data <- read.csv("Comparative_primate_steepness_values.csv")

#la phylogénie
specieslist<-as.data.frame(matrix(unique(data$species),ncol=1,nrow=length(unique(data$species))))
colnames(specieslist)<-"species"
rownames(specieslist)<-specieslist$species


phylogeny <- read.nexus("Upham2019MammalPhylogeny.nex")

# check whether the species in our data are the same as the species in the phylogeny
speciesmatching<-name.check(phylogeny,specieslist)
# check which species from our data are not in the phylogeny
speciesmatching$data_not_tree

specieslist<-as.data.frame(matrix(unique(data$species),ncol=1,nrow=length(unique(data$species))))
colnames(specieslist)<-"species"
rownames(specieslist)<-specieslist$species

# and reduce the phylogeny to only include the species for which we have data
speciesmatching<-name.check(phylogeny,specieslist)
mtree<-drop.tip(phylogeny,speciesmatching$tree_not_data)
```

Preparing the data
```{r}
#Nombre de matrices par espèce
data_N <- data %>%
  group_by(species) %>%
  summarise(Nmatrices = n(), .groups = "drop") %>%
  mutate(
    species_label = str_replace_all(species, "_", " "),
    species_label = str_to_sentence(species_label)
  )

#steepness moyenes par espèce et par sexe
data_steep <- data %>%
  group_by(species, sex) %>%
  summarise(
    mean_steepness = mean(steepness, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    sex = recode(sex,
                 "M" = "males",
                 "F" = "females",
                 "male" = "males",
                 "female" = "females",
                 "males" = "males",
                 "females" = "females")
  )

data_steep_wide <- data_steep %>%
  pivot_wider(
    names_from = sex,
    values_from = mean_steepness
  )


```

Initial tree
```{r}
p <- ggtree(mtree) + theme_tree2()
p2 <- p %<+% data_N

#etirement vertical
y_expand <- 2
p2$data$y <- p2$data$y * y_expand
```

Extract tips positions
```{r}
tip_positions <- p2$data %>%
  filter(isTip) %>%
  select(label, x, y) %>%
  rename(species = label)

tree_max_x <- max(tip_positions$x)
```

Build values positions 
```{r}
text_data <- data_steep_wide %>%
  left_join(tip_positions, by = "species") %>%
  mutate(
    x_anchor  = tree_max_x + 27,
    x_male    = x_anchor,
    x_female  = x_anchor + 4.8
  )
```

legend
```{r}
clade_legend <- data.frame(
  clade = factor(c("Platyrrhini", "Cercopithecoidea", "Hominoidea"),
                 levels = c("Platyrrhini", "Cercopithecoidea", "Hominoidea")),
  x = NA, y = NA
)

sex_legend <- data.frame(
  sex = factor(c("males", "females"), levels = c("males", "females")),
  x = 1,
  y = 1,
  label = "0.88"
)
```

plot
```{r}
p_final <- p2 +
  geom_hilight(node = 70, fill = "#8da0cb", alpha = 0.18) +
  geom_hilight(node = 41, fill = "#66c2a5", alpha = 0.18) +
  geom_hilight(node = 67, fill = "#fc8d62", alpha = 0.18) +

  geom_tippoint(aes(size = Nmatrices),
                shape = 21, fill = "steelblue", alpha = 0.7) +

  geom_tiplab(aes(label = species_label),
              size = 3.8,
              offset = 4,
              align = TRUE,
              linetype = "dotted",
              fontface = "italic") +

 # valeurs mâles
geom_text(
  data = (text_data %>% filter(!is.na(males)) %>% mutate(sex = "males")),
  aes(x = x_male, y = y, label = sprintf("%.2f", males), color = sex),
  inherit.aes = FALSE,
  hjust = 0,
  size = 3.5,
  fontface = "bold"
) +

# valeurs femelles
geom_text(
  data = (text_data %>% filter(!is.na(females)) %>% mutate(sex = "females")),
  aes(x = x_female, y = y, label = sprintf("%.2f", females), color = sex),
  inherit.aes = FALSE,
  hjust = 0,
  size = 3.5,
  fontface = "bold"
) +

  # titre unique
  annotate(
    "text",
    x = text_data$x_anchor[1] + 3,
    y = max(text_data$y) + 3.5,
    label = "Average steepness",
    fontface = "bold",
    size = 4
  ) +

  # légende clade
  geom_point(
    data = clade_legend,
    aes(x = x, y = y, fill = clade),
    shape = 22,
    size = 4,
    inherit.aes = FALSE
  ) +
  
  scale_fill_manual(
    values = c(
      "Platyrrhini" = "#8da0cb",
      "Cercopithecoidea" = "#66c2a5",
      "Hominoidea" = "#fc8d62"
    ),
    name = "Clade"
  ) +
  
  scale_color_manual(
    values = c("males" = "#FDE725FF", "females" = "#443A83FF"),
    name = NULL
  ) +

  scale_size_area(
    max_size = 6,
    name = "N matrices",
    breaks = round(seq(min(data_N$Nmatrices),
                       max(data_N$Nmatrices),
                       length.out = 4))
  ) +

  guides(
    fill = guide_legend(order = 1, override.aes = list(shape = 22, size = 5)),
    color = guide_legend(
      override.aes = list(
        label = c("0.88", "0.88"),
        shape = NA
      )
    ),
    size = guide_legend(order = 3)
  ) +

  coord_cartesian(
    xlim = c(0, max(text_data$x_female) + 5),
    clip = "off"
  ) +
  theme_tree2() +
  theme(
    legend.position = "right",
    plot.margin = margin(20, 70, 20, 10)
  )

p_final
```













